RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-07-31 manuscripts/manuscript.Rmd
mention phylobase among further directions #83

There are many promising directions for further development of the
`RNeXML` package.  The project's [Milestones page](https://github.com/ropensci/RNeXML/issues/milestones) page provides a list of
steps currently planned or under consideration.  In particular, these
milestones include increased support for interactive mode of entering data
(wizards), and support for integration of semantics defined through the
Web Ontology language (OWL), as discussed in the Challenges section.

>>>>

There are many promising directions for further development of the
`RNeXML` package.  The project's [Milestones page](https://github.com/ropensci/RNeXML/issues/milestones) page provides a list of
steps currently planned or under consideration.  In particular, these
milestones include increased support for interactive mode of entering data
(wizards), conversions to and from other existing R phylogenetics formats
such those found in the `phylobase` package `r citep(citation("phylobase"))`,
and support for integration of semantics defined through the
Web Ontology language (OWL), as discussed in the Challenges section.

### 2014-07-31 manuscripts/manuscript.tex
bump version and add notes post CRAN-release
mention phylobase among further directions #83

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic
Data in R}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[NESCent]{Hilmar Lapp}

    \author[SimonFraser]{Scott Chamberlain}

@@

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[NESCent]{National Evolutionary Synthesis Center, Durham, NC, USA}    
    \address[SimonFraser]{Department of Biology, Simon Fraser University, Burnaby, Canada}    
    \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}

@@

The R language (R Core Team 2014) has become an important environment
for phylogenetics research, with over 30 packages available including
several with over 1000 citations (`CRAN task view' 2014), though
stand-alone programs remain an important part of this ecosystem for
tasks such as generating phylogenetic trees from sequence data. Within
this ecosystem a variety of different data formats for phylogenetic data
have been developed and modified. Most but not all are based on the
simple and popular \texttt{phylo} format provided by the ape package
(Paradis \emph{et al.} 2004).

@@

We begin by reading in an example NeXML file provided with the
package.\footnote{For convenience \texttt{nexml\_read} is also aliased
  as \texttt{read.nexml}, for users familiar with the somewhat analogous
  \texttt{read.nexus} function.}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{f <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"trees.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(f)}
\end{Highlighting}
\end{Shaded}

The resulting \texttt{nex} object is an R object representation of the
NeXML. This is a native R object for representing phylogenetic trees (as
well as character traits, and metadata), analogous to the \texttt{phylo}
object from the popular \texttt{ape} R package (Paradis \emph{et al.}
2004). While R functions and packages could be written to use this
\texttt{nexml} object directly, most users will prefer to use already
existing methods provided by packages they know. From \texttt{nex}
object we can therefore extract any phylogenies it contains in the
\texttt{ape::phylo} format:

@@

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting
function from the \texttt{ape} package to plot the resulting
\texttt{ape::phylo} object returned by \texttt{get\_trees}.

@@

NeXML can not only contain multiple phylogenetic trees in a single
document, but can also group sets of trees into separate blocks (e.g.~as
resulting from different workflows). The \texttt{get\_trees} function
attempts to return the most sensible R object for NeXML: if only a
single tree is available, it will return an \texttt{ape::phylo} object;
if multiple trees are available in a single \texttt{trees} block, it
will return an \texttt{ape::multiPhylo} object (a list of \texttt{phylo}
objects). If multiple blocks are available, it will return a list of
\texttt{multiPhylo} blocks. While most users will find this behavior
convenient, there are times (such as looping over many NeXML files) when
having a variable return type is undesirable. To always get back a list
of \texttt{multiPhylo} objects, even when only one tree is available,
use \texttt{get\_trees\_list()}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex =}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Like NEXUS files, NeXML documents can also contain character data. While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological)
data used to infer the phylogenetic tree, we find this ideal for
character data used in comparative methods popular in R. Methods to
extract character data work much like those for the phylogenetic data:

@@

Returns a \texttt{data.frame} with columns as characters and rows as
taxa. Continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by \texttt{get\_characters()} if they correspond to
the same taxa. The \texttt{get\_characters\_list()} function will
consistently receive a list of data frames corresponding to each
character matrix block of the NeXML.

@@

The first argument to \texttt{nexml\_write} can be a \texttt{phylo}
phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
object -- anything that can be coerced into the \texttt{nexml} class. As
we have seen, NeXML files can contain multiple phylogenetic trees, as
well as character matrices. We can add both trees and character data
using \texttt{nexml\_write},

@@

To adjust these and other common metadata fields that should be part of
any data deposition, \texttt{RNeXML} provides the
\texttt{add\_basic\_meta()} function. This function does not assume any
familiarity with the concepts of namespaces or other more advanced
metadata tools discussed later. The \texttt{add\_basic\_meta} takes an
existing \texttt{nexml} object (or will create a new one if none is
provided, like the other \texttt{add\_} functions), and any of the
following fields: \texttt{title}, \texttt{description},
\texttt{creator}, \texttt{pubdate}, \texttt{rights}, \texttt{publisher},
\texttt{citation}. Additional fields may be added to subsequent
versions, see \texttt{?add\_basic\_metadata} for the most up-to-date
information. Here we create an new \texttt{nexml} object with the basic
metadata for the \texttt{bird.orders} data provided in the \texttt{ape}
package:

@@

  \DataTypeTok{description =}\StringTok{" This data set describes the phylogenetic relationships of the}
\StringTok{     orders of birds as reported by Sibley and Ahlquist (1990). Sibley}
\StringTok{     and Ahlquist inferred this phylogeny from an extensive number of}
\StringTok{     DNA/DNA hybridization experiments. The ``tapestry'' reported by}
\StringTok{     these two authors (more than 1000 species out of the ca. 9000}
\StringTok{     extant bird species) generated a lot of debates.}

@@

or to papers by simply using the object's DOI (using Boettiger (2014)):

@@

(Which is the paper on which the \texttt{geospiza} phylogeny from the
geiger dataset is based)

@@

\texttt{RNeXML} provides various functions to assist in automatically
adding other more specific metadata to various elements of the NeXML
data. One common use case is that of taxonomic identifiers. Misspellings
or inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come
from other places that use different taxonomies. Hence, not just
misspellings, but also synonyms, homonyms and incompatible taxon
concepts might plague the end user. By associating the taxonomic labels
against a specific, centralized database we can catch some of these
inconsistencies and indicate more precisely which classification we are
appealing to in the event of conflicting taxonomic definitions. In this
example, we indicate that we intend the bird orders listed in the tree
to correspond to the NCBI definitions

@@

This function uses the \texttt{taxize} R library (Chamberlain \& Sz{ö}cs
2013) to check each taxon label against the NCBI database. If a unique
match is found, a metadata annotation is added to the taxon providing
the NCBI identification number to the taxonomic unit. If no match is
found, the user is warned to check for possible typographic errors in
the taxonomic labels provided.

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance. In order
to be machine-readable and understandable, NeXML requires all metadata
annotations follow the RDFa format (see Vos et al. (2012)) which
explicitly declares the meaning of the annotation in a way a computer
can understand. For instance, the title we gave to the
\texttt{bird.orders} phylogeny appears in the NeXML as:

@@

We see the title appears as the value of the \texttt{content} attribute
of this \texttt{meta} element. The \texttt{id} attribute provides us
with a unique way to refer to this meta element in this document. The
\texttt{datatype} tells us that the content is a text string (as opposed
to, say, a numerical or boolean value). The \texttt{property} is the
attribute that most interests us here; it tells us that the content is a
title. The prefix \texttt{dc} indicates the vocabulary (here identified
as an XML namespace) to which the word \texttt{title} belongs. The
prefix is defined elsewhere in the NeXML file, as we can see with:

@@

Of course writing out such a definition manually becomes tedious
quickly. Because these are just R commands, we can easily define a
function that can loop over an assignment like this for each edge,
extracting the appropriate order, length and state from an existing R
object such as that provided in the \texttt{phytools}
package.\\Likewise, it is straight forward to define a function that
reads this data using the \texttt{RNeXML} utilities and converts it back
to the \texttt{phytools} package. The full implementation of this
mapping can be seen in the \texttt{simmap\_to\_nexml()} and the
\texttt{nexml\_to\_simmap()} functions provided in the \texttt{RNeXML}
package.

@@

Using this convention we can generate NeXML files containing simmap
data, read those files into R, and convert them back into the
\texttt{phytools} package format. These simple functions serve as
further illustration of how \texttt{RNeXML} can be used to extend the
NeXML standard. We illustrate their use briefly here, starting with
loading a \texttt{nexml} object containing a simmap reconstruction into
R:

@@

The \texttt{get\_trees} function can be used to return an
\texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
the simmap reconstruction data and returns includes this in a
\texttt{maps} element of the \texttt{phylo} object, for use with other
\texttt{phytools} functions.

@@

We can then use various functions from \texttt{phytools} designed for
simmap objects (Revell 2012), such as the plotting function:

@@

Though other NeXML parsers (for instance, for Perl or Python) have not
been written explicitly to express simmap data, those parsers will
nonetheless be able to successfully parse this file and expose the
simmap data to the user.

@@

The \texttt{RNeXML} package is available through the CRAN repository
under a CC0 license. At the time of writing, the current version is
1.0-2. The current development branch can be found on Github at
{[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for
changes in recent versions. Users are encouraged to submit bug reports
or feature requests in the issues log. Contact
\texttt{ropensci-discuss@googlegroups.com} or the package maintainer for
help, or email the phylogenetics R users group list at
\texttt{r-sig-phylo@r-project.org}.

There are many promising directions for further development of the
\texttt{RNeXML} package. The project's {[}Milestones
page{]}{[}\url{https://github.com/ropensci/RNeXML/issues/milestones}{]}
page provides a list of steps currently planned or under consideration.
In particular, these milestones include increased support for
interactive mode of entering data (wizards), and support for integration
of semantics defined through the Web Ontology language (OWL), as
discussed in the Challenges section.

@@

This project was supported in part by a grant from the Alfred P Sloan
Foundation (CB and SC), NSF Grant DBI-1306697 (CB), \ldots{}

\texttt{RNeXML} is part of the rOpenSci project,
{[}\url{http://ropensci.org}{]}.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{RNeXML: a package for reading and writing richly annotated phylogenetic,
character, and trait data in R}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[NESCent]{Hilmar Lapp}

    \author[ropensci]{Scott Chamberlain}

@@

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[NESCent]{National Evolutionary Synthesis Center, Durham, NC, USA}    
    \address[ropensci]{University of California, Berkeley, CA, USA}    
    \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}

@@

The R language (R Core Team 2014) has become an important environment
for phylogenetics research, with over 30 packages available including
several with over 1000 citations (`CRAN task view' 2014), though
stand-alone programs remain an important part of this ecosystem for
tasks such as generating phylogenetic trees from sequence data. Within
this ecosystem a variety of different data formats for phylogenetic data
have been developed and modified. Most but not all are based on the
simple and popular \texttt{ape::phylo} format provided by the ape
package (Paradis \emph{et al.} 2004).

@@

We begin by loading the \texttt{RNeXML} package and reading in an
example NeXML file provided with the package, containing two small
phylogenetic trees.\footnote{For convenience \texttt{nexml\_read} is
  also aliased as \texttt{read.nexml}, for users familiar with the
  somewhat analogous \texttt{read.nexus} function.}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{f <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"trees.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(f)}
\end{Highlighting}
\end{Shaded}

The resulting \texttt{nex} object is an R object representation of the
NeXML. This is a native R object, of the \texttt{nexml} class, for
representing phylogenetic trees (as well as character traits, and
metadata), analogous to the \texttt{ape::phylo} object from the popular
\texttt{ape} R package (Paradis \emph{et al.} 2004). While R functions
and packages could be written to use this \texttt{nexml} object
directly, most users will prefer to use already existing methods
provided by packages they know. From a \texttt{nexml} object, we can
therefore extract any phylogenies it contains in the \texttt{ape::phylo}
format. In this case we will get back a \texttt{multiPhylo} object,
since the file contains two phylogenies.

@@

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting
function from the \texttt{ape} package to plot the first tree in the
resulting \texttt{ape::phylo} object returned by \texttt{get\_trees}.

@@

NeXML can not only contain multiple phylogenetic trees in a single
document, but can also group sets of trees into separate blocks (e.g.~as
resulting from different workflows). The \texttt{get\_trees} function
attempts to return the most sensible R object for NeXML: if only a
single tree is available, it will return an \texttt{ape::phylo} object;
if multiple trees are available in a single \texttt{trees} block, it
will return an \texttt{ape::multiPhylo} object (a list of
\texttt{ape::phylo} objects). If multiple blocks are available, it will
return a list of \texttt{multiPhylo} blocks. While most users will find
this behavior convenient, there are times (such as looping over many
NeXML files) when having a variable return type is undesirable. To
always get back a list of \texttt{multiPhylo} objects, even when only
one tree is available, use \texttt{get\_trees\_list()}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Like NEXUS files, NeXML documents can also contain character data. While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological)
data used to infer the phylogenetic tree, we find this ideal for
character data used in comparative methods popular in R. Methods to
extract character data work much like those for the phylogenetic data.
Here we read in an example NeXML file that contains both a phylogeny and
character data. \texttt{get\_characters} returns a \texttt{data.frame}
with columns as characters and rows as taxa:

@@

As this example shows, continuous and discrete characters are
represented as separate blocks in NeXML files, but will be combined as
separate columns of a single data frame by \texttt{get\_characters()} if
they correspond to the same taxa. The \texttt{get\_characters\_list()}
function will consistently receive a list of data frames corresponding
to each character matrix block of the NeXML.

@@

The first argument to \texttt{nexml\_write} can be a \texttt{ape::phylo}
phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
object -- anything that can be coerced into the \texttt{nexml} class. As
we have seen, NeXML files can contain multiple phylogenetic trees, as
well as character matrices. We can add both trees and character data
using \texttt{nexml\_write},

@@

To adjust these and other common metadata fields that should be part of
any data deposition, \texttt{RNeXML} provides the
\texttt{add\_basic\_meta()} function. This function does not assume any
familiarity with the concepts of namespaces or other more advanced
metadata tools discussed later. The \texttt{add\_basic\_meta} takes an
existing \texttt{nexml} object (or will create a new one if none is
provided, like the other \texttt{add\_} functions), and any of the
following fields: \texttt{title}, \texttt{description},
\texttt{creator}, \texttt{pubdate}, \texttt{rights}, \texttt{publisher},
\texttt{citation}. Additional fields may be added to subsequent
versions, see \texttt{?add\_basic\_metadata} for the most up-to-date
information. Here we create a new \texttt{nexml} object with the basic
metadata for the \texttt{bird.orders} data provided in the \texttt{ape}
package:

@@

  \DataTypeTok{description =} \StringTok{"This data set describes the phylogenetic relationships of the}
\StringTok{     orders of birds as reported by Sibley and Ahlquist (1990). Sibley}
\StringTok{     and Ahlquist inferred this phylogeny from an extensive number of}
\StringTok{     DNA/DNA hybridization experiments. The ``tapestry'' reported by}
\StringTok{     these two authors (more than 1000 species out of the ca. 9000}
\StringTok{     extant bird species) generated a lot of debates.}

@@

or to papers by simply using the object's DOI (using Boettiger (2014)).
For instance to add the citation information of the paper that generated
the \texttt{geospiza} phylogeny included in the \texttt{geiger} package:

@@

@@

\texttt{RNeXML} provides various functions to assist in automatically
adding other more specific metadata to various elements of the NeXML
data. One common use case is that of taxonomic identifiers. Misspellings
or inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come
from other places that use different taxonomies. Hence, not just
misspellings, but also synonyms, homonyms and incompatible taxon
concepts might plague the end user. By associating the taxonomic labels
against a specific, centralized database we can catch some of these
inconsistencies and indicate more precisely which classification we are
appealing to in the event of conflicting taxonomic definitions. In this
example, we indicate that we intend the bird orders listed in the tree
to correspond to the NCBI definitions.

The \texttt{taxize\_nexml} function uses the R package \texttt{taxize}
(Chamberlain \& Sz{ö}cs 2013) to check each taxon label against the NCBI
database. If a unique match is found, a metadata annotation is added to
the taxon providing the NCBI identification number to the taxonomic
unit. If no match is found, the user is warned to check for possible
typographic errors in the taxonomic labels provided.

@@

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance. In order
to be machine-readable and understandable, NeXML requires all metadata
annotations follow the RDFa format (see Vos et al. (2012)) which
explicitly declares the meaning of the annotation.

Some basic definitions about the XML syntax used to express NeXML will
be helpful for this discussion, though a more comprehensive introduction
can be found in Vos et al. (2012). For instance, consider how the title
we gave to the \texttt{bird.orders} phylogeny appears in the NeXML file
using the XML format:

@@

The format begins with an XML ``tag'' and the name ``meta'', indicated
by the opening angle bracket; \texttt{\textless{}meta}. A closing tag
\texttt{\textbackslash{}\textgreater{}} indicates the end of this XML
``element'' . This tag contains several ``attributes'', named ``id'',
``property'', etc., that an R user can think of as names in a
\texttt{list}. What follows the equals sign after the attribute name is
called the attribute ``value'', (e.g.~the \texttt{id} attribute has
value ``m2'').

The name of any element (like ``meta'') or attribute (like ``id'') in
XML must be associated with a particular dictionary called a
``namespace'' and specified using a ``prefix'' of an abbreviation
followed by a colon appearing before the name. If no prefix is given, as
is the case for all attributes here: \texttt{id}, \texttt{property},
\texttt{datatype}, and \texttt{content}, then a default prefix specified
at the beginning of the document applies, in this case, the NeXML
namespace. This allows users to combine different XML that have
different definitions of terms like \texttt{id} or \texttt{property}.

The example element above also uses namespaces on some of the attribute
values themselves. This is a consequence of RDFa annotation, and not
strictly part of XML, though it works much the same way. While we hope
the following examples will make these concepts clearer, users do not
need to understand XML and namespaces to use most of the basic
annotation features provided by \texttt{RNeXML}.

So much for the structure; we can now briefly explain the meaning of
this element. We see the title appears as the value of the
\texttt{content} attribute of this \texttt{meta} element. The
\texttt{id} attribute provides us with a unique way to refer to this
meta element in this document. The \texttt{datatype} tells us that the
content is a text string (as opposed to, say, a numerical or boolean
value). The \texttt{property} is the attribute that most interests us
here; it tells us that the content is a title. The prefix \texttt{dc}
indicates the vocabulary (here identified as an XML namespace) to which
the word \texttt{title} belongs. The prefix is defined elsewhere in the
NeXML file, as we can see with:

@@

Of course writing out such a definition manually becomes tedious
quickly. Because these are just R commands, we can easily define a
function that can loop over an assignment like this for each edge,
extracting the appropriate order, length and state from an existing R
object such as that provided in the \texttt{phytools}
package.\\Likewise, it is straightforward to define a function that
reads this data using the \texttt{RNeXML} utilities and converts it back
to the \texttt{phytools} package. The full implementation of this
mapping can be seen in the \texttt{simmap\_to\_nexml()} and the
\texttt{nexml\_to\_simmap()} functions provided in the \texttt{RNeXML}
package.

@@

Using this convention we can generate NeXML files containing
\texttt{simmap} data, read those files into R, and convert them back
into the \texttt{phytools} package format. These simple functions serve
as further illustration of how \texttt{RNeXML} can be used to extend the
NeXML standard. We illustrate their use briefly here, starting with
loading a \texttt{nexml} object containing a \texttt{simmap}
reconstruction into R:

@@

The \texttt{get\_trees} function can be used to return an
\texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
the \texttt{simmap} reconstruction data and returns includes this in a
\texttt{maps} element of the \texttt{ape::phylo} object, for use with
other \texttt{phytools} functions.

@@

We can then use various functions from \texttt{phytools} designed for
\texttt{simmap} objects (Revell 2012), such as the plotting function:

@@

Though other NeXML parsers (for instance, for Perl or Python) have not
been written explicitly to express \texttt{simmap} data, those parsers
will nonetheless be able to successfully parse this file and expose the
\texttt{simmap} data to the user.

@@

The \texttt{RNeXML} package is available through the CRAN repository
under a CC0 license. At the time of writing, the current version is
1.1.1. The current development branch can be found on Github at
{[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for
changes in recent versions. Users are encouraged to submit bug reports
or feature requests in the issues log. Contact
\texttt{ropensci-discuss@googlegroups.com} or the package maintainer for
help, or email the phylogenetics R users group list at
\texttt{r-sig-phylo@r-project.org}.

There are many promising directions for further development of the
\texttt{RNeXML} package. The project's
\href{https://github.com/ropensci/RNeXML/issues/milestones}{Milestones
page} page provides a list of steps currently planned or under
consideration. In particular, these milestones include increased support
for interactive mode of entering data (wizards), conversions to and from
other existing R phylogenetics formats such those found in the
\texttt{phylobase} package (al. 2014), and support for integration of
semantics defined through the Web Ontology language (OWL), as discussed
in the Challenges section.

@@

.

This project was supported in part by a grant from the Alfred P Sloan
Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB),
\ldots{} We thank F. Michonneau for helpful comments on an earlier
version of this manuscript. \texttt{RNeXML} is part of the rOpenSci
project, {[}\url{http://ropensci.org}{]}.

al., R.H. et. (2014). \emph{phylobase: Base package for phylogenetic
structures and comparative data}. Retrieved from
\url{http://CRAN.R-project.org/package=phylobase}
